---
title: "Visual Analysis of Hotel Bookings Dataset"
author: "Dominik Vukelic"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, fig.show='hold', message=FALSE)
## Importing necessary libraries
library(ggplot2)
library(tidyverse)  # Includes dplyr and forcats
library(scales)
library(ggrepel)

Sys.setlocale("LC_TIME", "English")
```

```{r}
# Load the dataset directly within the R code chunk with the full path
dataset_path <- "C:\\Users\\Hectic\\Documents\\hotel_bookings_dataset_project\\cleaned_starting_file\\hotel_bookings_cleaned.csv"

df <- read.csv(dataset_path, na.strings = c("", "NA"))
```

# Creating a pie chart with percentages for cancelations

```{r}

df_percent <- df %>%
  count(is_canceled) %>%
  mutate(percentage = n / sum(n) * 100)


ggplot(df_percent, aes(x = "", y = percentage, fill = is_canceled)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(theta = "y") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_stack(vjust = 0.5)) +
  labs(title = "Cancellation Status",
       fill = "Cancellation Status") +
  scale_fill_manual(values = c("yes" = "lightpink", "no" = "lightblue")) +
  theme_void()
```

# Creating a grouped Bar Plot for Cancellation Counts by Hotel Type

```{r}

ggplot(data = df, aes(x = hotel_type, fill = factor(is_canceled))) +
  geom_bar(position = "dodge", color = "black") +
  labs(title = "Cancellation Counts by Hotel Type", y = "Count", fill = "Canceled or not") +
  scale_fill_manual(values = c("yes" = "maroon", "no" = "skyblue"))
```

# Creating a plot for Booking Occurrence Over Time

```{r}

df$arrival_date <- as.Date(df$arrival_date)



ggplot(data = df, aes(x = arrival_date)) +
  geom_line(stat = "count", color = "blue") +  
  labs(title = "Booking Occurrence Over Time", x = "Booking Date", y = "Number of Bookings") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  theme_minimal()  +  
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  
```

# Creating a histogram for stay durations

```{r}

ggplot(data = df, aes(x = total_stay)) +
  geom_histogram(binwidth = 1, fill = "limegreen", color = "black") +
  labs(title = "Distribution of Stay Duration", x = "Stay Duration (days)", y = "Frequency")
```

# Creating a bar plot for meal type

```{r}
ggplot(data = df, aes(x = meal, fill = factor(meal))) +
  geom_bar(position = position_dodge(width = 0.9), stat = "count", color = "black") +  
  geom_text(stat = "count", aes(label = stat(count)), position = position_dodge(width = 0.9), vjust = -0.5) +  
  labs(title = "Meal Type Preference", x = "", y = "") +
  scale_fill_discrete(name = "Meal Type") +  
  theme(axis.text.x = element_blank(), axis.text.y = element_blank(),  
        axis.title.x = element_blank(), axis.title.y = element_blank(), 
        axis.ticks = element_blank()) 
```

# Creating a histogram for stay durations with separate facets for weekend and weekday

```{r}

df_long <- pivot_longer(df, cols = c(stays_in_weekend_nights, stays_in_week_nights), names_to = "Stay_Type", values_to = "Stay_Duration")


ggplot(data = df_long, aes(x = Stay_Duration, fill = Stay_Type)) +
  geom_histogram(binwidth = 1, position = "dodge", color = "black") +
  labs(title = "Distribution of Stay Duration", x = "Stay Duration (days)", y = "Frequency") +
  facet_wrap(~ Stay_Type, scales = "free_y", labeller = labeller(Stay_Type = c("stays_in_weekend_nights" = "Weekend Stay", "stays_in_week_nights" = "Week Stay"))) +
  scale_fill_manual(values = c("stays_in_weekend_nights" = "blue", "stays_in_week_nights" = "red"), name = "Stay Type")

```

# Creating a violin plot to compare stay duration for previous and non-previous guests

```{r}

ggplot(data = df, aes(x = previous_guest, y = total_stay, fill = previous_guest)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.2, fill = "white", color = "black", outlier.shape = NA) +
  labs(title = "Stay Duration vs. Previous Guest Status", x = "Previous Guest", y = "Stay Duration (days)")
```

# Creating a bar plot for booking distribution by month

```{r}

df_counts <- df %>%
  group_by(month = factor(month(arrival_date, label = TRUE)), hotel_type) %>%
  summarise(count = n())


ggplot(data = df_counts, aes(x = month, y = count, fill = hotel_type)) +
  geom_col(position = position_dodge(width = 1), color = "black", width = 0.9) +
  geom_text(aes(label = count, group = hotel_type),
            position = position_dodge(width = 1), vjust = -0.5, size = 3) + 
  labs(title = "Booking Distribution by Month", x = "Month", y = "Count", fill = "Hotel Type") +
  scale_x_discrete(labels = month.abb) 
```

# Creating a grouped bar plot for the top 10 countries showing the counts of each hotel type

```{r}

top_countries <- df %>%
  group_by(country) %>%
  summarise(total_count = n()) %>%
  arrange(desc(total_count)) %>%
  head(10)

filtered_data <- df %>%
  filter(country %in% top_countries$country)


ggplot(data = filtered_data, aes(x = country, fill = hotel_type)) +
  geom_bar(position = position_dodge(width = 0.8), color = "black", width = 0.7) +  
  geom_text(stat = "count", aes(label = ..count..), position = position_dodge(width = 0.8), vjust = -0.5, size = 3) +  
  labs(title = "Hotel Type Distribution in Top 10 Countries", x = "Country", y = "Count", fill = "Hotel Type") +
  theme_minimal() +  
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  
  scale_fill_brewer(palette = "Set2")  
```

# Creating a grouped bar chart for non-canceled reservations and guest types

```{r}

df <- df %>%
  mutate(guest_type = case_when(
    adults > 0 & children == 0 & babies == 0 ~ "Adults Only",
    adults > 0 & children > 0 & babies == 0 ~ "Adults with Children",
    adults > 0 & babies > 0 & children == 0 ~ "Adults with Babies",
    adults > 0 & children > 0 & babies > 0 ~ "Adults with Children and Babies",
    TRUE ~ "Other"
  ))


non_canceled_data <- df %>%
  filter(is_canceled == "no")


ggplot(data = non_canceled_data, aes(x = guest_type, fill = guest_type)) +
  geom_bar(color = "black", position = position_dodge(width = 0.9), width = 0.7) +  
  geom_text(stat = "count", aes(label = ..count..), position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +  
  labs(title = "Guest Type Distribution for Non-Canceled Reservations") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.title.y = element_blank()) +  
  scale_fill_brewer(palette = "Set2")  
```

# Creating a histogram for booking changes

```{r}

ggplot(data = df, aes(x = booking_changes)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 3) +  
  scale_x_continuous(breaks = unique(df$booking_changes)) +  
  labs(title = "Distribution of Booking Changes", x = "Booking Changes", y = "Frequency")
```

